---
title: "The Impact of Attitudes on the Gender Wage Gap"
subtitle: "My subtitle if needed"
author: 
  - Alaina Hu
thanks: "Code and data are available at: https://github.com/alainahu/gender_wage_gap"
date: today
date-format: long
abstract: "This study examines the impact of societal attitudes towards women's income, GDP, and education level on the gender wage gap. Through statistical analysis of data from various countries, we find that negative attitudes towards women's income significantly exacerbate the gender wage gap, notwithstanding the mitigating effects of higher education levels and GDP. Our findings highlight the critical role of societal attitudes in economic disparities between genders, suggesting targeted interventions for policymakers. This research underscores the complexity of the gender wage gap and the importance of addressing societal attitudes to achieve gender equality in the labor market."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```


# Introduction

The gender wage gap, the average difference in pay between men and women, persists as a prominent issue in labor economics, reflecting broader societal inequalities. Despite significant strides towards gender equality, disparities in earnings remain a stubborn challenge worldwide. The causes of the gender wage gap are multifaceted, involving complex interactions between economic, social, and educational factors. Traditional economic explanations such as differences in occupation, industry, work experience, and education have been extensively studied. However, these factors do not fully account for the persistent wage discrepancies, suggesting the influence of more intangible elements, such as societal attitudes towards women’s roles in the workforce and the economy.

Recent research has begun to explore the impact of these societal attitudes, proposing that they may significantly influence economic outcomes for women. This paper seeks to contribute to this emerging body of research by examining how attitudes towards women's income, alongside GDP and education level, predict variations in the gender wage gap across different contexts. Social psychology research has revealed the importance of attitudes on behavior. By integrating data on societal attitudes with economic and educational indicators, this study aims to uncover the nuanced ways in which these variables interact to affect the economic valuation of women's labor. The estimand is the impact of societal attitudes toward women's income, GDP, and education on the gender wage gap. 

Our research paper begins with the Data section (@sec-data) to visualize and further understand the measurement, source, methodology, and variables we are examining. Then, we introduce the Model (@sec-model) used to understand the relationships in the data and report the findings in the Results section (@sec-results). Finally, we include the Discussion (@sec-discussion) of the findings, summarizing the takeaway and future of this research.



# Data {#sec-data}

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it. 



# Model {#sec-model}

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.  

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.


### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.


# Results {#sec-results}

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```




# Discussion {#sec-discussion}

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this. 

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}


# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows... 

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```



\newpage


# References


